Last glacial cycle glacier erosion potential in the Alps

### 2020-12-28 articles/erosion/alps-erosion.tex
One sentence per paragraph in erosion paper.

    The glacial landscape of the Alps has fascinated generations of explorers,
    artists, mountaineers and scientists for centuries.

    In fact, the global obsession for the Alps has been so large that in
    English, a non-native language foreign to the Alps, the word “alpine” is
    now used to describe any glacially modified mountain landscape worldwide.

    Understanding the beauty of Alpine glacial landscape through the rigour of
    scientific research is yet another endeavour.

    While it has long been understood that glaciers leave a specific imprint on
    the landscape they lie on, observing and quantifying the long-term erosion
    and sedimentation processes has been a challenge. While some regions are
    specifically characterised by cirque glaciation (), or glacial scouring (),
    the European Alps present a variety of glacial landforms with a density
    seen in few other places on Earth.

    It is typically assumed that the degree of glacial modification is a proxy
    for the duration of past glacier cover. However glacier sliding velocities
    vary by several orders of magnitude, and erosion laws can be non-linear.
    Besides, cold-based glaciers have been shown to preserve landscape.

@@

@@

@@

@@

@@

@@

@@

@@

>>>>

    The glacial erosion landscape of the Alps has fascinated generations of
    explorers, artists, and scientists for centuries. In English, a non-Alpine
    language, the word “alpine” is now used to describe any glacially modified
    mountain landscape worldwide.

    Some regions are characterised by cirque glaciation, glacial valleys,
    or overland scouring. Other regions, including the Alps, present a variety
    of glacial erosional landforms. It is often assumed that the degree of
    glacial modification is a proxy for the duration of past glacier cover.

    While it has long been understood that glaciers leave a specific imprint on
    the landscape they lie on, observing and quantifying the long-term erosion
    and sedimentation processes has been a challenge. Glacier sliding
    velocities vary by several orders of magnitude, and erosion laws can be
    non-linear. Besides, cold-based glaciers have been shown to preserve
    landscape, but it remains unclear how sharp the erosional transition to
    cold-base ice is, and why some glaciated regions show traces of cold-base
    glaciations and others not.

    Here we apply a non-linear erosion law to previously published model
    results and analyse the patterns of modelled erosion rate and cumulative
    last glacial cycle erosion potential.

@@

    Summary of PISM physics and boundary conditions.

@@

    We follow the near-quadratic erosion law of \citep{Herman.etal.2015}.

    We refer to erosion potential as the time-integrated erosion rate.

@@

    Modelled cumulative erosion potential vary by several orders of magnitude.
    There is a very strong localization in valleys with a large catchment and
    valley heads, where the results are limited by model resolution.

@@

    Above a certain level of glaciation, erosion rates appear uncorrelated with
    the ice volume. Volumetric erosion rates during MIS~4 and MIS~2 are not
    higher than other parts of the last glacial cycle. However, there is a
    tendency for higher erosion rate during periods of ice retreat.

    Along the Rhine Glacier transect, stages of extensive glaciation see
    erosion rates localized in the lower part of the catchment. Erosion rates
    propagate inward during periods of ice retreat.

@@

    These results can be generalized to the entire model domain by using
    elevation as a proxy for along-flow distance. During periods of larger
    extent, erosion rates shift to lower elevation, leaving upper grounds
    untouched. During deglaciation, erosion propagates inland.

@@

% -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -
\subsection{Results interpretation}
% -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -

    It is often believed that glaciers are more erosive during deglaciations
    because of higher water availability. However, surface meltwater is absent
    from the model. Higher erosion rates can be explained by enhanced ice
    dynamics after longer-term ice cover.

    The results also indicate that if a square-rule erosion law is valid, much
    of the intra-Alpine erosional landscape do not date from stages of major
    glaciation.

    The modelled erosion potential depends significantly on the choice of
    pleoclimate forcing applied, but the general patterns remain similar.
    Higher precipitation yields higher erosion. The results also appear to be
    sensitive to model resolution, primarily in the mountains interior, where
    discretized slopes depend on the grid size.

@@

    The cumulative erosion potential modelled by the Franz-Joseph Glacier
    erosion law \citep{Herman.etal.2015} appear overestimated. However, the
    error bars are large.

    The erosion law by \citet{Cook.etal.2020} results in similarly high values
    but a flatter (less localized) pattern of cumulative erosion potential,
    with lower values in the troughs and more erosion in the mountains. The two
    erosion laws may appear incompatible, and there would be a lot of reasons
    for them to be incompatible. But as pointed by \citet{Cook.etal.2020}, they
    are not necessarily incompatible. The erosion's feedback onto ice dynamics
    may yield to a delocalization of high-velocity flow in the long run.

    The erosion law by \citet{Koppes.etal.2015} yields a similar pattern but
    lower values. This law is for tidewater glaciers, which flow much faster
    than mountain glaciers. However, we can imagine that Alpine paleoglaciers
    flowing on a thick sediment bed (and sometimes into proglacial lakes, not
    included), behave closer to tidewater glaciers than the Franz-Josef steep
    mountain glacier.

@@

    The modelled erosion rates are very highly uncertain. The results are
    limited as there is no feedback of erosion onto ice dynamics, and
    erosion by rivers is not included. But if a square-law is used:
    \begin{itemize}
      \item Cumulative erosion potential is strongly localized in regions
        of fast past glacier flow.
      \item The total erosion rate is not correlated with the ice volume.
      \item During major glaciations, erosion rates are high at low elevation,
        but the mountain's interior is preserved.
      \item Higher-elevation glacial erosional landforms were formed during
        stages of intermediate glaciation.
      \item Erosion increases during deglaciations, regardless of surface
        meltwater availability.
    \end{itemize}
